\label{sec:solver-evidence}

This section collects supporting solver evidence that is methodologically useful
but too implementation-specific for the main benchmark narratives.

\subsection{Plasticity3D Source-Assembly and Fixed-Operator PMG Comparisons}

The primary Plasticity3D numerical story in \Cref{sec:num-plasticity3d} is the
maintained local path. The tables below add two engineering comparisons on the
same discrete problem from \cref{eq:plasticity3d-discrete}: a matched
source-assembly comparator and a fixed source-operator PMG variant.

\begin{table}[H]
  \centering
  \input{tables/generated/plasticity3d_local_vs_source.tex}
  \caption{Matched local-vs-source comparison for the same local PMG solver
  stack. At each listed rank the local and source rows converge in the same
  number of nonlinear and linear iterations, so this is a timing-only
  comparison on \cref{eq:plasticity3d-discrete}.}
  \label{tab:plasticity3d-local-vs-source}
\end{table}

\begin{table}[H]
  \centering
  \input{tables/generated/plasticity3d_fixed_source_operator_pmg.tex}
  \caption{Fixed source-operator PMG alternative on the same Plasticity3D
  problem. The status column distinguishes completed rows from rank counts
  that hit the Newton cap before reaching the promoted gradient target.}
  \label{tab:plasticity3d-fixed-source-pmg}
\end{table}

These tables support two narrower statements. First, source assembly has lower
reported wall time than maintained local assembly when both are embedded in the
same PMG solver stack, although the gap narrows at higher rank counts. Second,
the fixed source-operator PMG profile is not the preferred maintained default in the
present study: it may appear favorable on some rows, but it does more nonlinear
and Krylov work and converges less consistently across the reported runs.

\subsection{Source continuation evidence}

The legacy source-continuation path is not the mainline method of the paper,
but it is still useful because it isolates how much preconditioner policy
matters even when the outer continuation logic is held fixed.

\begin{table}[H]
  \centering
  \input{tables/generated/source_continuation_compare.tex}
  \caption{Source continuation with the fixed PMG-shell smoother policy at 8
  and 32 ranks. This is supporting solver evidence only; it is not part of the
  paper's main maintained production story.}
  \label{tab:source-cont}
\end{table}

\subsection{Code, data, and reproducibility availability}

The implementation discussed in this paper is available at
\url{https://github.com/Beremi/fenics_nonlinear_energies}. The manuscript is
generated from the repository-local paper pipeline. The table, figure,
literature-source, asset-validation, and reproducibility-note scripts live in
\path{paper/scripts}. The generated manuscript assets live in
\path{paper/tables/generated} and \path{paper/figures/generated}. The raw
benchmark summaries used by those scripts are kept under
\path{artifacts/raw_results}. The paper build validates that the generated
assets are present before compiling the PDF. A citable source or artifact
archive DOI should be supplied in the final submission metadata if required by
the target journal; this journal-neutral draft does not assert one.
