This paper presented \repo{} as a software-and-methods workflow for nonlinear
finite-element solves and optimization with a maintained \jaxpetsc{} production
path. The central contribution is not a new constitutive theory or a universal
differentiable-PDE abstraction, but a maintained implementation pattern in
which local JAX differentiation, sparse PETSc nonlinear solves, and multiple
second-order routes are kept together across a heterogeneous benchmark suite.

The Plasticity3D evidence is intentionally narrower than a broad mechanics
claim. Same-case source-faithfulness to the source-style surrogate
implementation is strong, and the fixed-lambda source-operator comparison
passes the configured endpoint field-agreement checks after matching the
glued-bottom boundary contract. These results support the maintained
endpoint-surrogate workflow for this benchmark family, while still not
supporting constitutive-equivalence language or path-consistent
incremental-history claims. Constitutive AD emerges as the preferred
cost-quality route on the locked flagship case. On Hyperelasticity, the
companion comparison
against JAX-FEM shows that the maintained code can agree very closely with a
recent external implementation when the mesh and prescribed displacement
schedule are matched and terminal states are post-compared with the repository
energy. The promoted Plasticity3D timing study and the distributed topology
workflow then show that the same codebase remains operational beyond small
validation comparisons.

Future work should follow the boundaries made visible by these results:
\begin{itemize}
  \item extend the validation surface for path-dependent mechanics beyond the
        present surrogate-versus-source-operator comparison, including a true
        incremental-history reference;
  \item add more fairness-first external baselines on tightly matched problem
        contracts;
  \item continue reducing linear-setup and multigrid costs in the flagship 3D
        mechanics workflows.
\end{itemize}

The next scientific step is not to overclaim this implementation, but
to turn this maintained workflow into a stronger reusable template for
additional nonlinear multiphysics and design problems with equally explicit
validation and fairness criteria.
